---
title:  "ADA2: Class 13, Ch 08, polynomial regression"
author: Sina Mokhtar
date:   last-modified
description: |
  [Advanced Data Analysis 2](https://StatAcumen.com/teach/ada12,
  Stat 428/528, Spring 2023, Prof. Erik Erhardt, UNM
format:
  html:
    theme:            litera
    highlight-style:  atom-one
    page-layout:      full      # article, full   # https://quarto.org/docs/output-formats/page-layout.html
    toc:              true
    toc-location:     body  # body, left, right
    number-sections:  false
    self-contained:         false     # !!! this can cause a render error
    code-overflow:          scroll    # scroll, wrap
    code-block-bg:          true
    code-block-border-left: "#30B0E0"
    code-copy:              false     # true, false, hover a copy buttom in top-right of code block
fig-width:                  6
fig-height:                 4
fig-align:                  center    # default, left, right, or center
execute: # https://quarto.org/docs/computations/execution-options.html, https://quarto.org/docs/computations/r.html
  cache:    false   # false, true
  echo:     true    # true, false  Include the source code in output
  warning:  true    # true, false  Include warnings in the output.
  error:    true    # true, false  Include errors in the output (note that this implies that errors executing code will not halt processing of the document).
---

# Hooker's Himalayian boiling point altitude data

Dr. Joseph Hooker collected the following data in the 1840s on the boiling
point of water and the atmospheric pressure at 31 locations in the Himalayas.
Boiling point is measured in degrees Fahrenheit. The pressure is recorded in
inches of mercury, adjusted for the difference between the ambient air
temperature when he took the measurements and a standard temperature.

The goal was to develop a model to predict the atmospheric pressure from the
boiling point.

**Historical note:** Hooker really wanted to estimate altitude above sea level from
measurements of the boiling point of water. He knew that the altitude could be
determined from the atmospheric pressure, measured with a barometer, with lower
pressures corresponding to higher altitudes. His interest in the above
modelling problem was motivated by the difficulty of transporting the fragile
barometers of the 1840s. Measuring the boiling point would give travelers a
quick way to estimate elevation, using the known relationship between elevation
and barometric pressure, and the above model relating pressure to boiling
point.

```{R}
library(erikmisc)
library(tidyverse)

dat_boil <-
  read_csv(
    "ADA2_CL_13_boilingpressure.csv"
  , skip = 2
  ) %>%
  mutate(
    boilingF_cen = boilingF - mean(boilingF)
  )

# x-variable mean for centering
dat_boil$boilingF %>% mean()

str(dat_boil)
```

## __(2 p)__ Plot the data.

Using `ggplot`, try to implement these features in a plot.
Overlay both a straight-line regression line in blue (`geom_smooth(method = lm, col = "blue", ...)`),
as well as a loess smooth (default) dashed line in red (`geom_smooth(method = loess, col = "red", linetype = 2, ...)`).
Using `alpha=1/5` will make the confidence bands more transparent.
Also, if you plot the points last, they'll lie on top of the lines.

Describe the key features of this plot.

### Solution

I'll give you this first plot to help get started, in particular to illustrate a nice use of the caption
  and the annotation of a second x-axis for the centered version of the `boilingF` variable.

```{R, fig.height = 5, fig.width = 5}
library(ggplot2)
p <- ggplot(dat_boil, aes(x = boilingF, y = pressure))
p <- p + scale_x_continuous(sec.axis = sec_axis(~ . - mean(dat_boil$boilingF), name = "boilingF centered"))
p <- p + geom_vline(xintercept = mean(dat_boil$boilingF), alpha = 1/4)
p <- p + geom_smooth(method = lm, se = TRUE, col = "blue", fill = "blue", alpha = 1/5)
p <- p + geom_smooth(method = loess, se = TRUE, col = "red", fill = "red", linetype = 2, alpha = 1/5)
p <- p + geom_point(size = 2)
p <- p + labs(title = "Simple linear model"
            , caption = "Blue solid = line, Red dashed = loess smooth curve"
              )
print(p)
```


[answer]


## __(3 p)__ Fit a simple linear regression, assess assumptions.

Fit a simple linear regression model for predicting pressure from boiling
point. Provide output for examining residuals, outliers, and influential cases.

Looking at the plots, are there any indications that the mean pressure is not
linearly related to boiling point? Are there any observations that appear to be
highly influencing the fit of this model? Are there certain points or regions
of the data where the model does not appear to fit well? Discuss.

Which, if any, of the standard linear regression model assumptions appears to
be violated in this analysis? If you believe that some of the assumptions are
violated, does it appear that deleting one or two points would dramatically
improve the fit? Would you use this model for predicting pressure from boiling
point? Discuss and carry out any needed analysis to support your position.

### Solution

[answer]
```{R}
names(dat_boil)
model.lm = lm(
  pressure ~ boilingF,
  data=dat_boil
  )
summary(model.lm)
```

```{r fig.height = 5,  fig.width =5}
e_plot_lm_diagostics(model.lm)

```

there is curve in qqplot but its inside the bound so its roughly normally distributed. the observation 1, 2, 3 has lot influence in compare to others but its mostly because of the lack of fit. the observation 1 is highly influential and also large leverage but overall the whole model is not fit. The residuals vs fitted value and residual vs boilingF plot is curvature which mean pressure is not linearly related to boiling point and our model does not fit with the data. the boxcox plot also  suggest log transformation. however this model do a precise prediction but we are not suggest simple linear model for this data, because lack of fit. overall its a good model but not the best.  





## __(1 p)__ Interpret $R^2$

Interpret $R^2$ in the previous simple linear regression model.

### Solution

[answer]

The $R^2$ in the previous model was `r summary(model.lm)$r.square` which means our model explain %99 variation in response so the model do a good job on prediction, however as we saw before the linear model does not fit the data so $R^2 $ does not tell anything about  fitting of a model.  




## __(2 p)__ A better model.

Decide whether transformation, or a polynomial model in boiling point, is
needed to adequately summarize the relationship between pressure and boiling
point. If so, perform a complete analysis of the data on this scale (that is,
check for influential observations, outliers, non-normality, etc.).

### Solution

[answer]
we have few options: we could fit a quadratic model or we can do log transformation (boxcox suggestion) we will try both to find a better model.



#### Quad model

```{R}
model.quad = lm(
  pressure ~ boilingF + I(boilingF^2),
  data=dat_boil
  )
summary(model.quad)
```

```{r fig.height = 5,  fig.width =5}
e_plot_lm_diagostics(model.quad)

```
In qqplot the residuals follow the line much better. the observation 2, 3 has more influence in compare to others but not as much as observation 1 in linear model. the observation 1 is highly influential and also large leverage. there is no structure in residual vs fitted value and residuals vs. boilingF and residuals vs boilingF^2. in boxcoxplot the peak is exactly on 1 which means we could resolve the issue of normality by adding boilingF^2 to the model. the added variable plots both look acceptable. overall all assumptions are met.

the boilingF^2 is significant and $R^2$ increase from`r summary(model.lm)$r.square`  to `r summary(model.quad)$r.square`. therefore this is an acceptable model. 

Note:
because the boilingF is not centered the only H0 test we can interpret here is the highest ordered term which is boilingF^2 however we can make a new model with centered data. all the residuals and plot and coefficient of boilingP^2 variable will be the same however the coef and p-value of boilingF variable will change. 





```{R}
model.quad.centered = lm(
  pressure ~ boilingF_cen + I(boilingF_cen^2),
  data=dat_boil
  )
summary(model.quad.centered)
```

```{r fig.height = 5,  fig.width =5}
e_plot_lm_diagostics(model.quad.centered)

```


#### log model
```{R}

model.log = lm(
  log(pressure) ~ boilingF,
  data=dat_boil
  )
summary(model.log)
```


```{r fig.height = 5,  fig.width =5}
e_plot_lm_diagostics(model.log)

```


In qqplot the residuals follow the line. the observation 2, 31 has more influence in compare to others. the observation 2 is influential and also large has large leverage. there is no structure in residual vs fitted value and residuals vs. boilingF and residuals vs boilingF^2 however is not as good as the quadratic model. in boxcoxplot the one is in acceptable interval. overall it was improved in compareed to simple linear model but not as good as quadratic model. 

the boilingF is significant and $R^2$ increase from`r summary(model.lm)$r.square` for linear model  to `r summary(model.log)$r.square`. in compare to $R^2$ : `r summary(model.quad)$r.square` for quadratic model it is little bit lower therefore this is an acceptable model but not as good as quadratic model. 


## __(2 p)__ Final model.

Regardless of which scale you choose for the analysis, provide an equation to
predict pressure from boiling point. Write a short summary, pointing out any
limitations of your analysis.

### Solution

[answer]
presure = 19.7 + 0.417 (boilF - mean(boilF)) + 0.00403 (boilF - mean(boilF))^2

We did fit a model with a very high R^2 value and also all assumptions have been met however The regression model only apply to the domain of the data that were sampled. there may be also some error in measurement because of fragile barometers that were been used. 


#### Example based on the first linear model

_Assuming you called your linear model object `model.quad`,
  then the equation with code below will place
  the intercept and slope in the equation.
Just add an `r ` before each of the `signif(...)` inline code chunks to
  make the numbers appear.
Then use this example to write your final model here._

$$
  \widehat{\textrm{pressure}}
  =
    `r signif(coef(model.quad.centered)[1], 3)`
  + `r signif(coef(model.quad.centered)[2], 3)` \textrm{ boilingF_cen}
  
$$


